### 2021-06-16 letter/responseLetter.tex

\vspace{0.2cm}

\item In the second place, I would recommend to include more details on the experiments conducted. This is especially 
true for the second study, where Section 3.3 is a bit shallow. The three steps of the methodology are indeed shortly 
described without providing enough details, hence possibly precluding verifiability and replicability. In addition, the 
methodology is not linked to the research questions: I would recommend to restructure Section 3 in order to have 
data extraction processes and data analysis procedures for each research question.

\vspace{0.2cm}

{\color{blue}{\bf Answer.} We restructured Section 3.3 to detail our approach better. We created sub Section 3.2.1 to present data extraction process of our First study, and 3.2.2 to present data analysis procedure. In some way, we created sub Section 3.3.1 and 3.3.2, to address the same issues, referring to our second study. With this restructuring we better clarify our study setting Section, presenting more details about our first (Section 3.2) and second study (Section 3.3). We also created a picture that can help to understand the methodology of our second study (Figure 3). }

@@

  \vspace{0.2cm}

At Section 1 (Introduction) we highlight the existence of these 2 repositories that can be useful for other researchers interested in our study.}

\vspace{0.2cm}

\item Another point to improve is connected to the implications of the study. While the paper discusses in nice way the 
results achieved, it does not provide a clear view of how those results can be potentially useful for researchers 
working on Android sandboxes as well as what are the next steps that the research community should pursue to 
improve the detection of malware.

@@

\vspace{0.2cm}

{\color{blue}{\bf Answer.} We agree. To mitigate this issue, we include at Section 6 (Conclusion and future work Section) one last paragraph that points how our results can be useful for researches at Android sandboxes. In the same paragraph, we also highlight which next steps can contribute to research evolution.}

@@

\begin{itemize}

\item The abstract is not clear. First, the context if not well addressed. For example, why ``mining Android sandboxes'' is 
important? I had to read several related reference at the beginning to understand it. Second, it is not common practice to 
include reference in abstract. However, I reserve my opinion since I know it may be preferred by some researchers.

\vspace{0.2cm}

{\color{blue}{\bf Answer.} We improved the abstract, making it clearer. We better contextualized ``mining sandbox approach'', presenting its definition e importance for Android malware analysis.}

@@

\vspace{0.2cm}

\item Also, FlowDroid requires as input the sources and sinks APIs. But you mine ``all sources and sinks 
paths''. Do you mean all APIs? FlowDroid will never finish running.

\vspace{0.2cm}

{\color{blue}{\bf Answer.} You are right: we actually use a curate set of sensitive sources and sinks
  that is available in the FlowDroid repository.
  We changed the first paragraph of Section 3.3.1 to clarify this point. Thanks.
  We hope that the current version of our paper fixes this problem.}

@@

\begin{itemize}

\item It is not clear how this paper can be considered an extension of the DroidXP paper presented in SCAM 2020.

\vspace{0.2cm}

{\color{blue}{\bf Answer.} In the introduction, we rephrase paragraphs 4 to better clarify how our paper is an extension of DroidXP paper from SCAM 2020.}

@@

\vspace{0.2cm}

\item ``The authors state that dynamic analysis exceeds static analysis on mining Android sandboxes because the
analysis must often assume that additional behaviors are possible than actually would be.''
Please explain

\vspace{0.2cm}

{\color{blue}{\bf Answer.} We explained better the sentence at Introduction, paragraph 2.}

@@

\vspace{0.2cm}

\item ``We presented and evaluated DroidXP in a conference paper [11]''
What is DroidXP? It should be introduced to a reader who is not familiar.

\vspace{0.2cm}

{\color{blue}{\bf Answer.} We agree with that. To mitigate this issue, we introduced better DroidXP at introduction, paragraph 4.}

@@

\vspace{0.2cm}

\item What are sensitive behaviors? ``showed that DroidBot is able to trigger an amount of sensitive behaviors''

\vspace{0.2cm}

{\color{blue}{\bf Answer.} Sensitive behaviors at Android context occurs when an Android app functionality can result in accessing or leaking of users sensitive data. Example of sensitive behaviors: File accesses and network access. We improve the text with this definition at Section 2 (Background and related works), paragraph 6.}

@@

\vspace{0.2cm}

\item Please rephrase:
``From the same authors, another test generator tool for Android, described as Humanoid [23], is a DroidBot evolution and
presents a proposal that can generate humans like tests inputs, using deep learning.''

\vspace{0.2cm}

{\color{blue}{\bf Answer.} We rephrase the text at Section 2 (Background and related works) paragraph 6, to better explain Humanoid test generation tool.}

@@

\vspace{0.2cm}

\item Briefly explain the taint analysis
``Taint analysis is a special type of static analysis that enables tracking of sensitive data within programs''

\vspace{0.2cm}

{\color{blue}{\bf Answer.} We agree that taint analysis explanation was very briefly. We reviewed our Section 2 (Background and related works), to better explain taint analysis from paragraph 8. We explained better the concept of ``source'' and ``sink'', and present how the analysis identifies sensitive information leakage, detecting taint flow between “source” and “sink”. To improve understanding, we also present a simple Listing (Listing 1), that present a simple data leakage example through data flows within source code Java. }

@@

\vspace{0.2cm}

\item Please explain:
``FastDroid [28] mitigates this issue by introducing an intermediate light-weight abstraction to perform the analysis.''

\vspace{0.2cm}

{\color{blue}{\bf Answer.} To address this issue, we explained better how FastDroid improves the taint analysis efficiency at the last paragraph of Section 2 (Background and related works)}

@@

\vspace{0.2cm}

\item Section 3.1, the help text is not adding much value

\vspace{0.2cm}

{\color{blue}{\bf Answer.} Thanks for raising this point. The section 3.1 presented our Benchmark tool DroidXP. We understand that the tool description was important for the paper since it was helpful for our first study conduction. DroidXP helped us to easily integrate the different test generation tools, and understand the effect of the static analysis component at each tool. It was useful for us and can be very useful for other researchers. We
reviewed the first paragraph of section 3.1 to clarify the importance of DroidXP for our study conduction.}

### 2021-06-16 conclusionfutureworks.tex
update all paper and response letter

In this paper, we performed two studies. In the first we conducted a replication of \blls, considering $96$ pairs apps (B/M) from the original study, and selecting $5$ test generation tools to run for $3$ minute for $3$ times. Firstly, we performed the study, using DroidXP benchmark with its default configuration, enabling the Droidfax static analysis algorithms, and repeat the study by disabling Droidfax, to compute the effective performance of the test generation tools, ignoring the influence of static analysis. For the second experiment, we complement the dynamic analysis using tainted analysis algorithms, through Flowdroid tool, over the same $96$ pairs (B/M) of the previous experiment. The results of the first experiment reveals that, ignoring our fake test case generation tool (Joker), that does not executed the apps, $58.33\%$-$76.04\%$ of the malware investigated was detected using the sandboxes with the support of Droidfax static analysis algorithms. The impact of Droidfax in the results range from $16.44\%$ (DroidBot) to $51.79\%$ (Humanoid). We also found that $18.75\%$ of malicious apps (18 out of 96) could be detected by none of the sandboxes generated. In the second experiment, the tainted analysis detected 58 among 96 pairs investigated (60.42\%). With the exception of DroidBot tool, tainted analysis algorithms had a better performance than any sandbox constructed by the execution of another test case tool, without the Droidfax static analysis algorithms. Our work show that the number of malicious behaviours detected is increased for all cases with tainted analysis aid, achieving at the best case, $82.29\%$ (79 among 96) of malware detected with DroidBot.

As future work, we plan to investigate other promising tools to enhance our data from previous executions. We will use the benchmark to test tools like Sapienz (a search-based test generation tool)~\cite{DBLP:conf/issta/MaoHJ16} and Dynodroid (an input generation system for Android apps) \cite{DBLP:conf/sigsoft/MachiryTN13}. And we will search for other great tools used by academia and by the industry to be tested by the benchmark DroidXP.
Also, we will execute each of the tools used in this experiment without the benchmark. And, we will compare these results with this experiment's results. This comparison will be valuable in the pursuit of one of this study's goals, which was to analyze what impact the benchmark droidXP has on its results.

>>>>

In this paper, we performed two studies. In the first we conducted a replication of \blls, considering $96$ pairs apps (B/M) from the original study, and selecting $5$ test generation tools to run for $3$ minute for $3$ times. Firstly, we performed the study, using DroidXP benchmark with its default configuration, enabling the Droidfax static analysis algorithms, and repeat the study by disabling Droidfax, to compute the effective performance of the test generation tools, ignoring the influence of static analysis. For the second experiment, we complement the dynamic analysis using taint analysis algorithms, through Flowdroid tool, over the same $96$ pairs (B/M) of the previous experiment. The results of the first experiment reveals that, ignoring our fake test case generation tool (Joker), that does not executed the apps, $58.33\%$-$76.04\%$ of the malware investigated was detected using the sandboxes with the support of Droidfax static analysis algorithms. The impact of Droidfax in the results range from $16.44\%$ (DroidBot) to $51.79\%$ (Humanoid). We also found that $18.75\%$ of malicious apps (18 out of 96) could be detected by none of the sandboxes generated. In the second experiment, the taint analysis detected 58 among 96 pairs investigated (60.42\%). With the exception of DroidBot tool, taint analysis algorithms had a better performance than any sandbox constructed by the execution of another test case tool, without the Droidfax static analysis algorithms. Our work show that the number of malicious behaviours detected is increased for all cases with taint analysis aid, achieving at the best case, $82.29\%$ (79 among 96) of malware detected with DroidBot.

As future work, we plan to investigate other promising tools to enhance our data from previous executions. We will use the benchmark to test tools like Sapienz (a search-based test generation tool)~\cite{DBLP:conf/issta/MaoHJ16} and Dynodroid (an input generation system for Android apps) \cite{DBLP:conf/sigsoft/MachiryTN13}. And we will search for other great tools used by academia and by the industry to be tested by the benchmark DroidXP.
Also, we will execute each of the tools used in this experiment without the benchmark. And, we will compare these results with this experiment's results. This comparison will be valuable in the pursuit of one of this study's goals, which was to analyze what impact the benchmark DroidXP has on its results.

Our research topic is very critical around the world, and contributions to this domain are necessary and should be considered strategic from the privacy perspective. The present work offered relevant insights into the role of dynamic and static Analysis for Android malware detection, and can be potentially useful for security researchers at Android platform. Our empirical findings demonstrate that static analysis techniques when combined with dynamic analysis at mining sandbox approach is a promising line of work
to improve the task of identifying malicious behavior at Android apps. As a next step, the research community should pursue explore possible static analysis techniques aimed at information flow analysis \cite{DBLP:conf/kbse/ShenVTADLKZ14} in general, and taint analysis \cite{arzt:pldi-2014} specifically, that we believe are possible to be integrated with dynamic analysis techniques, to compose a new mining proposal in sandbox. Another research line should also concentrate on the study of improvement in automatic test generation to modeling sandbox. Some studies point that code coverage of automatic test generation is not high \cite{DBLP:conf/sigsoft/ZengLZXDLYX16}\cite{DBLP:conf/kbse/ChoudharyGO15}, and needs improvement. When we are talking about ``improvement'', we refer not only to the ability to detect bugs but rather coverage of ``typical'' behavior, answering as accurately as possible the question: ``What an Android app actually proposes to do ?''

### 2021-06-16 setup.tex
update all paper and response letter

Answering the research questions RQ1 and RQ2 allows us to expose a possible overestimation in terms of the number of detected malware by each sandbox, constructed by the test
generation tools explored in \blls. Ignoring the implications of DroidFax in the \blls might have introduced a possible threat to their conclusions. Answering the third research question
allows us to open up the possibility of finding new strategies for designing mining sandbox techniques, complementing the performance of
dynamic analysis through the use of static analysis algorithms.
We conducted two studies to answer the research questions above. In the
first study we leverage DroidXP~\cite{DBLP:conf/scam/CostaMCMVBC20} (Section~\ref{sec:droidxp}) to replicate
the \blls. We present the settings of this first study in Section~\ref{sec:set1}.
In the second study we use
FlowDroid~\cite{DBLP:conf/pldi/ArztRFBBKTOM14} to investigate the 
suitability of taint analysis algorithms for mining Android sandboxes.
We present the settings of the second study in Section~\ref{sec:set2}.

@@

The \blls reports the results of an empirical study that compares the performance of test generation tools to mine Android
sandboxes, based on the calls to sensitive APIs from 102 pairs (benign / malign) of Android apps. In their work, \emph{performance} is
measured as the total number of malwares a sandbox is able to identify when comparing the calls to sensitive APIs made by
the benign and malign versions of the app. A \emph{hit} occurs whenever a sandbox finds that the malicious version of an app calls
additional sensitive APIs, in comparison to the calls from the corresponding benign version.
Nonetheless, the \blls uses DroidFax to instrument the Android apps, even though DroidFax also
collects calls to sensitive APIs using static analysis algorithms. Since the \blls does not
compute the possible impact of DroidFax into the performance of the test generation tools,
here we replicate their work to understand the impact of the DroidFax static analysis algorithms into the \blls results.

Our replication differs from the original work in a few decisions. First, here we isolate
the effect of the DroidFax static analysis algorithms, in the task to identify malicious apps. In addition, although we use the same dataset of
$102$ pairs of Android apps used in the \blls, here we discarded $6$ pairs for which
we were not able to instrument---out of the $102$ pairs used in the original work. We also introduced a recent test generator tool (Humanoid ~\cite{DBLP:conf/kbse/LiY0C19}), which
has not been considered in the previous work. Finally, we extended the execution time of each test generation tool,
executing each app from the test generation tool for three minutes (instead of one minute in the
original work),
and built the sandboxes after executing each test generation tool
three times---the original work executed each test generation tool
only once. That is, our goal here is not to conduct an
exact replication of the \blls, but instead understand
the role of the DroidFax static analysis algorithms in the
performance of test case generation tools for mining sandboxes.
% The original study executed each app at each tool, for just one minute, and just one time.

@@

% In the first study we executed the DroidXP benchmark with its
% default configuration, that is, enabling the DroidFax
% static analysis algorithms and the test generation tools.

We first run the experiment using the default
settings of DroidXP, to explore the four test case generation tools described earlier and a fake test
case generation tool (named \joke). \joke simulates a test tool that does not execute
the Android apps during a benchmark execution. Using this tool, the results
of the dynamic analysis are not considered and we can compute the results with
only the static analysis component of DroidFax (RQ1). Our study executed each pair of
Android app (B/M) in each one of the five test generation tools, including \joke,
for three minutes, and for three times. We then investigated the performance of each tool to detect malicious behaviors in the sandbox
modeled by the test generation tool
under analysis.

After that, we replicated the analysis using a configuration of DroidXP that disables the DroidFax static analysis algorithm.
Using this new configuration, we executed the benchmark again, using the same dataset of Android apps, the same execution time,
and the same test case generation tools.
The results of this replication allow us to compute the effective performance
of the dynamic analysis tools (RQ2)---that is, ignoring the influence of the
DroidFax static analysis algorithms.
Figure \ref{fig:setup} shows all the phases of our experiments, that explore all sensitive
resources access by both apps (B/M), and the optional configuration DroidXP, that disables DroidFax static analysis.

@@

In the second study 
we investigate whether or not a taint-based static analysis approach is also promising for
identifying malwares, given a version of an app that we can assume to be secure.
To this end, we leverage the FlowDroid
taint analysis algorithms for Android apps, in order to identify dataflows
that might lead to the leakage of sensitive information. Our
goal here is to investigate if it is possible to detect malicious
behavior by means of identifying the \emph{divergent} source-sink paths that FlowDroid reveals after
analysing a benign and a malign versions of an Android app.
define.

FlowDroid takes as input an Android Application Package (APK file) and
a set of API methods marked either as {\bf source}
or {\bf sink} (or both). Source methods are those that access \emph{sensitive information} (e.g.,
a method that access the address book), while sink methods are those 
that \emph{might share information with external peers} (e.g., a method that
sends messages to a recipient). We rely on the source-sink definitions
of the FlowDroid implementation~\cite{arzt:pldi-2014,rasthofer-source-sink},
which involves a curate list of source and sink methods (including callbacks and
other Android API methods of interest).
FlowDroid then uses a \emph{context, flow, and field
sensitive analysis} to identify all dataflow paths from sources to sinks~\cite{arzt:pldi-2014}.
However, it is important to note that not all source-sink paths are malicious, and then we
follow a specific methodology to identify malwares in our work: we raise a
warning (a \emph{hit}) whenever
FlowDroid finds an additional source-sink path in the malicious version of an app, which
has not been identified when analysing the benign version.

Our research settings involve three steps (see Figure~\ref{fig:settings2}).
In the first, we execute FlowDroid to mine the source-sink paths from a
benign version of an app, and then enumerate a set (S1) with the 
possible dataflows between sources and sinks. All paths in S1 are considered secure in
our analysis. In the second step we repeat the FlowDroid execution, though
considering the malicious APK version of the app.
This leads to a second set (S2) of source-sink paths. In the final step, we compute the difference (S3) between
the sets S2 and S1 (i.e., $S3 = S2 \setminus S1$), in order to reveal
the existence of additional paths in the
malicious version. If the set S3 is not empty, we conclude that FlowDroid
identified the malware.

We use two metrics in this second study:
the total number of malicious apps FlowDroid is able to find and the
execution time for running the taint analysis algorithm for each app.
Moreover, in this second study we use the same dataset of $96$ pairs of Android apps (B/M),
originally shared in the AndroZoo repository~\cite{DBLP:conf/msr/AllixBKT16} and
also used in the \blls.

>>>>

Answering the research questions RQ1 and RQ2 allows us to expose a possible overestimation in terms of the number of detected malware by each sandbox, constructed by the test
generation tools explored in \blls. Ignoring the implications of DroidFax in the \blls might have introduced a possible threat to their conclusions. Answering research question RQ3
allows us to open up the possibility of finding new strategies for malware detection, complementing the performance of
dynamic analysis through the use of static analysis algorithms.
We conducted two studies to answer the research questions above. In the
first study we leverage DroidXP~\cite{DBLP:conf/scam/CostaMCMVBC20} (Section~\ref{sec:droidxp}) to replicate
the \blls. We present the settings of this first study in Section~\ref{sec:set1}.
In the second study we use
FlowDroid~\cite{DBLP:conf/pldi/ArztRFBBKTOM14} to investigate the 
suitability of taint analysis algorithms to complement mining Android sandboxes. The main idea of this second study is explore the possibility of detected extra malware, do not detected at our first study, through another static analysis strategy based on taint analysis. We present the settings of the second study in Section~\ref{sec:set2}.

@@

The \blls reports the results of an empirical study that compares the performance of test generation tools to mine Android
sandboxes. Since the \blls does not
compute the possible impact of DroidFax into the performance of the test generation tools,
here we replicate their work to understand the impact of the DroidFax static analysis algorithms into the \blls results.

Our replication differs from the original work in a few decisions. First, here we isolate
the effect of the DroidFax static analysis algorithms, in the task to identify malicious apps. In addition, although we use the same dataset of
$102$ pairs of Android apps used in the \blls, here we discarded $6$ pairs for which
we were not able to instrument---out of the $102$ pairs used in the original work, originally shared in the AndroZoo repository~\cite{DBLP:conf/msr/AllixBKT16}. We also introduced a recent test generator tool (Humanoid ~\cite{DBLP:conf/kbse/LiY0C19}), which
has not been considered in the previous work. Finally, we extended the execution time of each test generation tool,
executing each app from the test generation tool for three minutes (instead of one minute in the
original work),
and built the sandboxes after executing each test generation tool
three times---the original work executed each test generation tool
only once. That is, our goal here is not to conduct an
exact replication of the \blls, but instead understand
the role of the DroidFax static analysis algorithms in the
performance of test case generation tools for mining sandboxes.
% The original study executed each app at each tool, for just one minute, and just one time.

@@

\subsubsection{Sensitive APIs extraction process}

In our first study, sensitive APIs extraction process starts with monitoring of invocations of sensitive APIs during the apps (already instrumented) execution by test generation tools under analysis. This task is essential for the process since sensitive APIs are used to access resource sensitives and it call could be a clue of private data leakage. Our study based at a list of 97 sensitive APIs, defined by the AppGuard privacy-control framework~\cite{DBLP:conf/esorics/BackesGHMS13}. Initially, we performances the extraction process using the default settings of DroidXP, (e.g., using DroidFax static analysis component). We explored the four test case generation tools described earlier and a fake test case generation tool (named \joke).

\joke simulates a test tool that does not execute the Android apps during the exploratory phase. Using this tool the results of the dynamic analysis are not considered, and we can compute the results with only the static analysis component of DroidFax (RQ1). Our study executed each 96 pair of
Android app (B/M) in each test generation tool, including \joke, for three minutes, and for three times.

Then we repeated this process, however this time we used a DroidXP configuration that disables the DroidFax static analysis algorithm.
With this new configuration, we monitored sensitive APIs invocations again, using the same 96 pair of Android app (B/M), the same execution time,
and the same test case generation tools. With this new configuration, became possible computer the effective performance of each test generation tool, in terms of malware detected, discarding the contributions of DroidFax static analysis algorithm (RQ2).

\subsubsection{Sensitive APIs analysis procedures}

After exploring all sensitive APIs by all test tools, we analyzed the \emph{performance} of these tools to detect malicious behaviors. The procedure starts with the analysis of each sandbox modeled by their respective test generation tools, based on the calls to sensitive APIs.

The test generation tool \emph{performance} is measured as the total number of malware, your respective sandbox is able to identify when comparing the calls to sensitive APIs made by the benign and malign versions of the app. We raise a warning (a \emph{hit}) whenever a sandbox finds that the malicious version of an app calls additional sensitive APIs, in comparison to the calls from the corresponding benign version. Figure \ref{fig:setup} shows all the phases of our experiments, that explore all sensitive resources access by both apps (B/M), and the optional configuration DroidXP, which disables DroidFax static analysis.

Assuming that our fake test case generation tool does not run the Android apps, its \emph{performance} analysis will reveal the DroidFax static algorithms impact at \blls, answering our first research question (RQ1). The analysis of the calls to sensitive APIs made by the 96 pair (B/M), when we disable the DroidFax static analysis component, will reveal the effective \emph{performance} of each test generation test tool under analysis, answering our second research question (RQ2).

@@

In the second study 
we investigate whether or not a taint-based static analysis approach is also promising for
identifying malwares, given a version of an app that we can assume to be secure.
To this end, we leverage the FlowDroid
taint analysis algorithms for Android apps, at version 2.8, in order to identify dataflows
that might lead to the leakage of sensitive information. Our
goal here is to investigate if it is possible to detect malicious
behavior by means of identifying the \emph{divergent} source-sink paths that FlowDroid reveals after
analysing a benign and a malign versions of an Android app.

\subsubsection{Source-sink paths extraction}

FlowDroid takes as input an Android Application Package (APK file) and
a set of API methods marked either as {\bf source}
or {\bf sink} (or both). Source methods are those that access \emph{sensitive information} (e.g.,
a method that access the address book), while sink methods are those 
that \emph{might share information with external peers} (e.g., a method that
sends messages to a recipient). We rely on the source-sink definitions
of the FlowDroid implementation~\cite{arzt:pldi-2014,rasthofer-source-sink},
which involves a curate list of source and sink methods (including callbacks and
other Android API methods of interest).
FlowDroid then uses a \emph{context, flow, and field
sensitive analysis} to identify all dataflow paths from sources to sinks~\cite{arzt:pldi-2014}.

Source-sink paths extraction involve two steps (see Figure~\ref{fig:settings2}). In the first, we execute FlowDroid to mine the source-sink paths from a benign version of an app, and then enumerate a set (S1) with the 
possible dataflows between sources and sinks. All paths in S1 are considered secure for
our future analysis. In the second step we repeat the FlowDroid execution, though
considering the malicious APK version of the app.
This leads to a second set (S2) of source-sink paths.

\subsubsection{Source-sink paths analysis}

For our analysis it is important to note that not all source-sink paths are malicious, and then we
follow a specific methodology to identify malware in our work: we raise a warning (a \emph{hit}) whenever
FlowDroid finds an additional source-sink path in the malicious version of an app, which
has not been identified when analysing the benign version. For this end, we compute the difference (S3) between the sets S2 and S1 (i.e., $S3 = S2 \setminus S1$), in order to reveal the existence of additional paths in the malicious version (third step). If the set S3 is not empty, we conclude that FlowDroid
identified the malware.

We use two metrics in this second study:
the total number of malicious apps FlowDroid is able to find and the execution time for running the taint analysis algorithm for each app. Moreover, in this second study we use the same dataset of $96$ pairs of Android apps (B/M) used in the First study. Our second study will lead to discover what is the capacity of malware detection of taint analysis algorithms compared to mining sandbox approach, and its potential to complement dynamic analysis for mining sandboxes, thus answering our last research question (RQ3).
